Phenotypic evolution is more constrained in simpler food webs

### 2019-01-22 manuscript.tex
finalize manuscript for bioRxiv

\emph{Manuscript elements}: table 1, figure 1, figure 2, figure 3,
figure 4. All figures should be printed in color.

@@

\emph{Number of words}: 5075

@@

We conducted our study within a four-year old common garden experiment
of coastal willow (\emph{Salix hookeriana}) located at Humboldt Bay
National Wildlife Refuge (HBNWR) (40 40'53``N, 124 12'4''W) near Loleta,
California, USA. This common garden consists of 26 different willow
genotypes that were collected from a single population of willows
growing around Humboldt Bay. Stem cuttings of each genotype (25
replicates per genotypes) were planted in a completely randomized design
in two hectares of a former cattle pasture at HBNWR. Willows at our
study site begin flowering in February and reach their peak growth in
early August. During this study, willows had reached 5 - 9m in height.
Further details on the genotyping and planting of the common garden are
available in \citet{Barbour2015}.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work \citep{Barbour2016}
and other work with gall midges \citep{Weis1983, Heath2018}. First, we
measured gall diameter as the size of each gall chamber to the nearest
0.01 mm at its maximum diameter (perpendicular to the direction of plant
tissue growth). Our previous work has shown that larger galls are
associated with higher survival \citep{Barbour2016}. Second, we measured
clutch size by counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this by randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple of features of our study system suggest that larval densities may
be a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

@@

Our analyses consisted of three parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
phenotypic traits (\(z_i\)) of individuals \textemdash in each food-web
treatment. Second, we translated selection surfaces into the scale of
relative fitness (\(w\)) in order to calculate selection gradients.
Third, we used our estimates of selection gradients to characterize the
slope and curvature of the adaptive landscape. Note that inferring
adaptive landscapes from selection surfaces assumes that trait
distributions are multivariate normal \citep{Lande1983}. To approximate
this assumption, we log-transformed clutch size and square-root
transformed oviposition preference. We then scaled all phenotypic traits
(mean=0 and SD=1) in order to calculate standardized selection gradients
that were comparable across traits and with other studies of natural
selection.

@@

All analyses and visualizations were conducted in R \citep{R2018}.
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population \citep{Arnold2001}.
We also plot mean larval survival on a natural log scale to accurately
reflect the mathematical definition of the adaptive landscape
\citep{Arnold2003}.

@@

\indent We found that more gall midge traits experienced directional
selection in the simple (3 of 3) vs.~complex food-web treatment (1 of
3)(table \ref{Table:Gradients}). For example, we observed directional
selection for smaller clutch sizes in the simple food web, but there was
no evidence of selection acting on this trait in the complex food web
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This
absence of selection appeared to be a result of conflicting selection
pressures imposed by each guild of parasitoids (fig.
\ref{fig:UV_Landscape}). Specifically, when we subset our data to focus
on differences between parasitoid guilds, we found that larval
parasitoids actually impose directional selection for larger clutch
sizes (\(\beta_{\text{Clutch}}\)= 0.13 {[}0.04, 0.24{]}). In the simple
food web, we also observed clear evidence of directional selection for
midges to avoid ovipositing on plants with high densities of
conspecifics (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}B); however, this relationship was weaker in
complex food webs (table \ref{Table:Gradients}). This was likely a
result of larval parasitoids imposing greater mortality on egg
parasitoids at high gall midge densities (see Selection on the extended
phenotype of egg parasitoids section), and thus a less than additive
effect on gall midges. Chamber diameter experienced positive directional
selection in both food-web treatments, but selection was 66\% higher in
complex food webs (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}). This was not due to any difference between egg
and larval parasitoids (fig. \ref{fig:UV_Landscape}), but was due to the
lower average survival in complex food webs (contrast -0.27 {[}-0.11,
-0.42{]}).

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in complex vs. simple food webs.}
\label{Table:Gradients}
\centering
\begin{tabular}{lccc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Complex} & \textbf{Simple} & \textbf{Contrast}  \\ 
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
0.34 [
0.22,
0.48] }& 
\textbf{
0.21 [
0.12,
0.31] }&

-0.14 [
-0.27,
0] \\

$\beta_{\text{Clutch}}$ & 
0.06 [
-0.05,
0.17] & 
\textbf{
-0.09 [
-0.17,
-0.01] }& 
\textbf{
-0.15 [
-0.29,
-0.03] }\\

@@

-0.03 [
-0.21,
0.15] \\

@@

-0.03 [
-0.27,
0.2] \\

@@

-0.06 [
-0.32,
0.2] \\

@@

0.02 [
-0.14,
0.18] & 
\textbf{
-0.32 [
-0.64,
-0.01] }\\

@@

-0.02 [
-0.17,
0.12] \\

@@

0.12 [
-0.05,
0.3] \\

@@

-0.03 [
-0.18,
0.12] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in complex vs.~simple food webs.
Each panel corresponds to a different combination of phenotypic traits:
clutch size and chamber diameter (A); clutch size and oviposition
preference (B); oviposition preference and chamber diameter (C). Arrows
represent mean estimates of directional selection gradients, while
contours represent predicted larval survival of the mean phenotype
(i.e., all traits = 0) in each food-web treatment. Notice that arrows
point more toward a corner of the adaptive landscape for each
combination of traits in the simple vs.~complex food-web treatment. This
indicates that trait integration (covariance) is more strongly favored
in simpler vs.~more complex food webs. Note that mean larval survival is
plotted on a natural log scale to reflect the mathematical definition of
the adaptive landscape.}
\end{figure}

@@

Gall midge preference influenced the probability of observing egg
parasitoids in complex vs.~simple food webs (contrast
\(\alpha_{\text{Pref}}\)= 1.03 {[}0.04, 2.31{]} ). Specifically, larval
parasitoids decreased the probability of observing egg parasitoids at
high gall midge densities (fig. \ref{fig:EggPtoid_Selection}). Note
however that this trait-fitness relationship did not translate into a
significant difference in selection (contrast \(\beta_{\text{Pref}}\)=
0.25 {[}-0.09, 0.64{]} ). This was likely because there was a tendency
(albeit not statistically signficant) for a lower probability of
egg-parasitoid survival in complex vs.~simple food webs (contrast=0.1
{[} -0.09, 0.32{]}), which masks this trait-fitness relationship. We
also observed two other interesting trends, although neither
significantly differed between treatments. Specifically, there was a
tendency for the impact of larval parasitoids to increase nonlinearly at
higher gall midge densities (fig. \ref{fig:EggPtoid_Selection}; complex
\(\alpha_{\text{Pref:Pref}}\)= -0.49 {[}-1.04, -0.13{]}; but contrast
\(\alpha_{\text{Pref:Pref}}\)= 0.46 {[}0.02, 1.07{]}). There was also a
tendency for egg parasitoids to obtain a refuge in large gall chambers
at high gall midge densities (complex \(\alpha_{\text{Pref:Diam}}\)=
0.25 {[}0.01, 0.54{]}; but contrast \(\alpha_{\text{Pref:Diam}}\)= -0.25
{[}-0.6, 0.05{]}).

@@

The generality of our results likely depends on the relative abundance
and functional differences between consumers in a community. For
example, if consumers do not differ from each other, then we do not
expect changes in food-web complexity to modify selective constraints.
Also, many consumers may be at too low of abundances to impose selection
on their resources. Rank abundance curves \citep{Preston1948} and the
disproportionate number of weak interactions in diverse communities
\citep{Paine1992} support this notion. When consumers are abundant
though, the effect of food-web complexity will depend on whether
different species impose conflicting selection pressures or select for
distinct traits. For example, parisitoids and birds impose conflicting
selection pressures on the size of galls induced by the fly
\emph{Eurosta solidaginis} \citep{Weis1985, Abrahamson1997, Start2016}.
Recent studies in this system have shown that decreases in the relative
abundance of birds, due to either small patch sizes \citep{Start2016} or
proximity to urban areas \citep{Start2018} causes a shift from neutral
to directional selection on gall size. On the other hand, different
consumers may impose selection on different traits, favoring trait
integration and increasing genetic constraints. Examples of this include
strong genetic covariances in plant resistance to different insect
herbivores \citep{Maddox1990, Wise2007, Wise2013}, although there are
also examples where these covariances are weak
\citep{Roche1997, Barbour2015}, or vary from year-to-year
\citep{Johnson2007}. We suggest that gaining predictive insight to the
evolutionary consequences of food-web disassembly requires an
understanding of the mechanisms governing the assembly of trophic
interactions.

>>>>

\emph{Manuscript elements}: table 1, table 2, figure 1, figure 2, figure
3, figure 4. All figures should be printed in color.

@@

\emph{Number of words}: 5069

@@

We conducted our study within a four-year old common garden experiment
of coastal willow (\emph{Salix hookeriana}) located at Humboldt Bay
National Wildlife Refuge (HBNWR) (40 40'53``N, 124 12'4''W) near Loleta,
California, USA. This common garden consists of 26 different willow
genotypes that were collected from a single population of willows
growing around Humboldt Bay. Stem cuttings of each genotype (25
replicates per genotype) were planted in a completely randomized design
in two hectares of a former cattle pasture at HBNWR. Willows at our
study site begin flowering in February and reach their peak growth in
early August. During this study, willows had reached 5 - 9m in height.
Further details on the genotyping and planting of the common garden are
available in \citet{Barbour2015}.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work \citep{Barbour2016}
and other work with gall midges \citep{Weis1983, Heath2018}. First, we
measured gall diameter as the size of each gall chamber to the nearest
0.01 mm at its maximum diameter (perpendicular to the direction of plant
tissue growth). Our previous work has shown that larger galls are
associated with higher survival \citep{Barbour2016}. Second, we measured
clutch size by counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this by randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple features of our study system suggest that larval densities may be
a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

@@

Our analysis consisted of three parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
phenotypic traits (\(z_i\)) of individuals \textemdash in each food-web
treatment. Second, we translated selection surfaces into the scale of
relative fitness (\(w\)) in order to calculate selection gradients.
Third, we used our estimates of selection gradients to characterize the
slope and curvature of the adaptive landscape. Note that inferring
adaptive landscapes from selection surfaces assumes that trait
distributions are multivariate normal \citep{Lande1983}. To approximate
this assumption, we log-transformed clutch size and square-root
transformed oviposition preference. We then scaled all phenotypic traits
(mean=0 and SD=1) in order to calculate standardized selection gradients
that were comparable across traits and with other studies of natural
selection.

@@

All analyses and visualizations were conducted in R \citep{R2018}.
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population \citep{Arnold2001}.
We also plot mean larval survival on a natural log scale to accurately
reflect the mathematical definition of the adaptive landscape
\citep{Arnold2003}. All data and code to reproduce the reported results
are publically available on GitHub
(\url{https://github.com/mabarbour/complexity_selection}) and have been
archived on Zenodo (\url{https://zenodo.org/badge/latestdoi/108833263}).

@@

\indent We found that more gall midge traits experienced directional
selection in the simple (3 of 3) vs.~complex food-web treatment (1 of
3)(table \ref{Table:Gradients}). For example, we observed directional
selection for smaller clutch sizes in the simple food web, but there was
no evidence of selection acting on this trait in the complex food web
(table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This
absence of selection appeared to be a result of conflicting selection
pressures imposed by each guild of parasitoids. Specifically, when we
subset our data to focus on differences between parasitoid guilds, we
found that larval parasitoids actually impose directional selection for
larger clutch sizes (\(\beta_{\text{Clutch}}\)= 0.13 {[}0.04, 0.24{]}).
In the simple food web, we also observed clear evidence of directional
selection for midges to avoid ovipositing on plants with high densities
of conspecifics (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}B); however, this relationship was weaker in
complex food webs (table \ref{Table:Gradients}). This was likely a
result of larval parasitoids imposing greater mortality on egg
parasitoids at high gall midge densities (see Selection on the extended
phenotype of egg parasitoids section), and thus a less than additive
effect on gall midges. Chamber diameter experienced positive directional
selection in both food-web treatments, but selection was 66\% higher in
complex food webs (table \ref{Table:Gradients}; fig.
\ref{fig:UV_Landscape}). This was not due to any difference between egg
and larval parasitoids (\(\beta_{\text{Diam}}\) contrast = 0.01 {[}
0.15, -0.11{]}), but was due to the lower average survival in complex
food webs (\(\bar W\) contrast -0.27 {[}-0.11, -0.42{]}).

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in complex vs. simple food webs.}
\label{Table:Gradients}
\centering
\begin{tabular}{lccc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Complex} & \textbf{Simple} & \textbf{Contrast}  \\ 
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
0.34 [
0.22,
0.48] }& 
\textbf{
0.21 [
0.12,
0.31] }& 
\textbf{
0.14 [
0.27,
0] }\\

$\beta_{\text{Clutch}}$ & 
0.06 [
-0.05,
0.17] & 
\textbf{
-0.09 [
-0.17,
-0.01] }& 
\textbf{
0.15 [
0.29,
0.03] }\\

@@

0.03 [
0.21,
-0.15] \\

@@

0.03 [
0.27,
-0.2] \\

@@

0.06 [
0.32,
-0.2] \\

@@

0.02 [
-0.14,
0.18] & 
\textbf{
0.32 [
0.64,
0.01] }\\

@@

0.02 [
0.17,
-0.12] \\

@@

-0.12 [
0.05,
-0.3] \\

@@

0.03 [
0.18,
-0.12] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

@@

\begin{figure}
\centering
\includegraphics{analyses/MV_landscapes.pdf}
\caption{\label{fig:MV_Landscape}Two dimensional view of adaptive
landscapes of gall midge phenotypes in complex vs.~simple food webs.
Each panel corresponds to a different combination of phenotypic traits:
clutch size and chamber diameter (A); clutch size and oviposition
preference (B); oviposition preference and chamber diameter (C). Arrows
represent mean estimates of directional selection gradients, while
contours represent predicted larval survival of the mean phenotype in
each food-web treatment. Notice that arrows point more toward a corner
of the adaptive landscape for each combination of traits in the simple
vs.~complex food-web treatment. This indicates that trait integration
(covariance) is more strongly favored in simpler vs.~more complex food
webs. Note that mean larval survival is plotted on a natural log scale
to reflect the mathematical definition of the adaptive landscape.}
\end{figure}

@@

Only gall midge preference influenced the probability of observing egg
parasitoids in complex vs.~simple food webs (table
\ref{Table:ExtendedGradients}). Specifically, the impact of larval
parasitoids increased nonlinearly with higher gall midge densities (fig.
\ref{fig:EggPtoid_Selection}).

\begin{table}[h]
\caption{Standardized selection gradients acting on the extended phenotype of egg parasitoids (\textit{Platygaster} sp.)}
\label{Table:ExtendedGradients}
\centering
\begin{tabular}{lc}
\\ 
\hline
\textbf{Selection gradient} & \textbf{Contrast = Complex - Simple}  \\ 
\hline
$\beta_{\text{Diam}}$ &

-0.03 [
0.25,
-0.3] \\

$\beta_{\text{Clutch}}$ &

0.07 [
0.38,
-0.26] \\

$\beta_{\text{Pref}}$ &

-0.25 [
0.09,
-0.64] \\

$\gamma_{\text{Diam,Diam}}$ &

-0.05 [
0.33,
-0.43] \\

$\gamma_{\text{Clutch,Clutch}}$ &

-0.21 [
0.26,
-0.68] \\

$\gamma_{\text{Pref,Pref}}$ &

\textbf{
-0.46 [
-0.02,
-1.07] }\\

$\gamma_{\text{Diam,Clutch}}$ &

0 [
0.27,
-0.29] \\

$\gamma_{\text{Diam,Pref}}$ &

0.25 [
0.6,
-0.05] \\

$\gamma_{\text{Clutch,Pref}}$ &

-0.18 [
0.12,
-0.52] \\ 
\hline
\end{tabular}
\bigskip{}
\\
{\footnotesize Note: Values in brackets represent 95\% confidence intervals. Bold values indicate that the 95\% CI does not overlap zero. $\beta_{\text{Diam}}$ has been adjusted for bias.}
\end{table}

@@

The generality of our results likely depends on the relative abundance
and functional differences between consumers in a community. For
example, if consumers do not differ from each other, then we do not
expect changes in food-web complexity to modify selective constraints.
Also, many consumers may be at too low of abundances to impose selection
on their resources. Rank abundance curves \citep{Preston1948} and the
disproportionate number of weak interactions in diverse communities
\citep{Paine1992} support this notion. When consumers are abundant
though, the effect of food-web complexity will depend on whether
different species impose conflicting selection pressures or select for
distinct traits. For example, parisitoids and birds impose conflicting
selection pressures on the size of galls induced by the fly
\emph{Eurosta solidaginis} \citep{Weis1985, Abrahamson1997}. Recent
studies in this system have shown that decreases in the relative
abundance of birds, due to either small patch sizes \citep{Start2016} or
proximity to urban areas \citep{Start2018} causes a shift from neutral
to directional selection on gall size. On the other hand, different
consumers may impose selection on different traits, favoring trait
integration and increasing genetic constraints. Examples of this include
strong genetic covariances in plant resistance to different insect
herbivores \citep{Maddox1990, Wise2007, Wise2013}, although there are
also examples where these covariances are weak
\citep{Roche1997, Barbour2015}, or vary from year-to-year
\citep{Johnson2007}. We suggest that gaining predictive insight to the
evolutionary consequences of food-web disassembly requires an
understanding of the mechanisms governing the assembly of trophic
interactions.
